---
title: "Using R as a BI tool"
author: "<span class='bigtext'>david.gohel @ ardata.fr</span>"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ['default', 'libs/addstyles.css']
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, dev.args=list(bg='transparent', pointsize=10), dev = "svg", 
                      fig.width = 5, fig.height = 4)
library(tidyverse)
library(knitr)
library(officer)
library(flextable)
library(antaDraft)
library(DT)
library(lubridate)
knit_print.flextable = render_flextable
knit_print.regulartable = render_flextable

theme_ <- theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# About me

&gt; 15 years of experience with data :

* statistical software edition
* CRM software edition
* retails
* climate business intelligence
* web analytics
* ... to be continued

Favorite tools:
<img src="img/Rlogo.svg" height="18px">&nbsp;
`r icon::fa("git")`&nbsp;
`r icon::fa("google")`&nbsp;
  
R packages: [officer](http://cran.r-project.org/package=officer) 
[flextable](http://cran.r-project.org/package=flextable) 
[mschart](http://cran.r-project.org/package=mschart) 
[rvg](http://cran.r-project.org/package=rvg) 
[ggiraph](http://cran.r-project.org/package=ggiraph) 
[gdtools](http://cran.r-project.org/package=gdtools) 


---
class: middle, inverse, center

# Business intelligence

> Using internal and external <span class="bigtext">data</span> to provide 
insights <span class="bigtext">to help decision</span> 

---
class: middle


<p class="sing">BI is consummed by people who are deciding, evaluating or communicating...</p>


---
class: middle


<p class="sing">BI is been used on demand and daily/weekly/monthly</p>


---
class: middle


<p class="sing">BI is ubiquitous</p>

---
class: middle


<p class="sing">Technically speaking, it is always about:</p>


* Structuring data
* Aggregations
* Transformations
* KPI calculation and optimization
* Reporting



---
class: middle, inverse, center

# BI can be made with R


---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
]

.right-column[

R can now read and write all data format I may need: 

```{r echo=FALSE}
tibble::tribble(
       ~Format,     ~Package,
   "CSV, TEXT",  "readr",
       "EXCEL",  "readxl",
         "SAS",  "haven",
        "SPSS",  "haven",
       "STATA",  "haven",
        "JSON",  "jsonlite",
        "HTML",  "rvest",
         "XML",  "xml2",
        "YAML",  "yaml"
  ) %>% regulartable() %>% fontsize(size = 16, part = "all") %>% 
  theme_booktabs() %>% autofit() 
```

]



---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
]
.right-column[

R can now connect to any database (SGBD, big data systems, ...) I may need to access...

The [R consortium](https://www.r-consortium.org), [RStudio team](https://www.rstudio.com) cleaned and improved databases support. 

```{r eval=FALSE}
library(DBI)
con <- dbConnect(RPostgres::Postgres(), dbname = 'db_name', host = 'host', 
  port = 5432, user = 'username', password = 'pwd')
con <- dbConnect(odbc::odbc(), driver = "PostgreSQL Driver", database = "db_name",
  uid = "username", pwd = "pwd", host = "host", port = 5432)

res <- dbGetQuery(con, "SELECT * FROM studies limit 5")
```

See [Databases using R](http://db.rstudio.com/)


]




---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
  ## data processing
]
.right-column[

[tidyverse](https://www.tidyverse.org/) and [data.table](https://CRAN.R-project.org/package=data.table) packages are covering 99% of necessary operations (join, aggregation, filtering, sorting, etc.).

[fuzzyjoin](https://cran.r-project.org/package=fuzzyjoin) and [tidytext](https://cran.r-project.org/package=tidytext) are easy enough to work with difficult textual data.

]



---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
  ## data processing
  ## checking
]
.right-column[


[lumberjack](http://cran.r-project.org/package=lumberjack/) is providing a logger system to monitor changes in data.

[validate](http://cran.r-project.org/package=validate/) is offering a set of function that will check if data are valid.

Of course, [tidyverse](https://www.tidyverse.org/) and [data.table](https://CRAN.R-project.org/package=data.table) are ubiquitous in this area also.

]


---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
  ## data processing
  ## checking
  ## exploration
]
.right-column[


[ggplot2](http://cran.r-project.org/package=ggplot2/) is providing an easy API to plot everything.

[tidyverse](https://www.tidyverse.org/) and [data.table](https://CRAN.R-project.org/package=data.table)...

[htmlwidgets for R - gallery](http://gallery.htmlwidgets.org): interactive visualisations

]





---
class: middle

## Does R fit to BI?

.left-column[
  ## IO
  ## data processing
  ## checking
  ## exploration
  ## reporting
]
.right-column[

### Static reporting 

[rmarkdown](http://rmarkdown.rstudio.com/) is allowing the production of HTML, PDF, MS Word, HTML dashboard, HTML static web sites. Package [officer](http://cran.r-project.org/package=officer) is offering several options to produce Word and PowerPoint files and package  [openxlsx](http://cran.r-project.org/package=openxlsx) brings a rich API to produce Excel files. 

### Interactive reporting 

[shiny](https://www.rstudio.com/products/shiny-server-pro/) fits very well for small and medium audiences.

[ShinyProxy](https://www.shinyproxy.io/) is also a serious candidate for shiny application.

[RStudio Connect](https://www.rstudio.com/products/connect/) is more than decent enterprise solution (shiny + static pages + scheduler + easy user managment). 

]



---
# Successful architectures 

> those I has experienced...


.left-column[
  ## R client
]
.right-column[

R fits on an USB key. 

Perfect for simple projects.

]




---
# Successful architectures 

> those I has experienced...

.left-column[
  ## R client
  ## opencpu
]
.right-column[

R as a service; an HTTP based server that can be called 
from any application able to manage HTTP protocol.

Highly secure

Pleases the IT

Requires experienced R coders to develop packages

]






---
# Successful architectures 

> those I has experienced...

.left-column[
  ## R client
  ## opencpu
  ## RStudio Connect
]
.right-column[

Easy and reliable.

User management, content management, scheduler, etc.

It is not open source

]




---
# Successful architectures 

> those I has experienced...

.left-column[
  ## R client
  ## opencpu
  ## RStudio Connect
  ## Rscript
]
.right-column[

Good old `crontab` or `Luigi` (build complex pipelines of batch jobs).

See also [littler](https://CRAN.R-project.org/package=littler).

Pleases the IT

For simple jobs

]



---
class: middle, inverse, center

# BI with R - pros and cons




---
# BI with R - The good

--

Thanks to its package ecosystem, R has <span class="bigtext">no functional limit</span>. 

--

By allowing pretty and innovative outputs, <span class="bigtext">R value our work</span>.

--

It cost only an R coder to try and test.

--

**Self BI is now possible and will be required more and more.**

--

**R community will help!**



---

# BI with R - Cons

--

Resistance from IT teams. *R is a live ecosystem, not a finite solution ready to be installed*.

--

Can be unoptimized. A newbie script is often just a proof it can be done, not a complete and robust analytic.

--

R is a glue system made of many packages and structured by its history, IT IS NOT a database system, it does not come with a visual GUI. 

--

13 000 packages! Choice issue... Again, need for a someone experimented with R.


---
class: middle, inverse, center

# Stories

---

## Reporting with packages `officer` and `mschart`


Story: a [yougov](http://www.yougov.com/) service whole role is to *do some BI* was spending too much time preparing PowerPoint documents for business analysts. 

Solution: [mschart](http://cran.r-project.org/package=mschart) to be able to deliver pre-formatted PowerPoints charts from an R analytic. (founder: yougov)

> business analysts still can annotate graphical reports and adapt content without leaving their favorite tool and without vampiring other services precious time.

```{r results='hide'}
library(officer)
library(mschart)
my_barchart <- ms_barchart(data = browser_data, x = "browser", y = "value", group = "serie")
doc <- read_pptx()
doc <- add_slide(doc, layout = "Title and Content", master = "Office Theme")
doc <- ph_with_chart(doc, chart = my_barchart)
print(doc, target = "samples/barchart_example.pptx")
```

[barchart_example.pptx](samples/barchart_example.pptx)

---
## Clinical reporting

Story: reduce time to update clinical reporting (Word format): 

* minimum 80 pages of clinical tables and graphics
* CDISC model
* This has to be made several times per study, there are several studies... There is a limited amount of time in a team of 6.

Solution: Developped packages for clinical studies treatment and reporting. When clinical database is updated, reports are generated.


---
## Web analytics stories

Story: a department is not willing to share its data internally, BI coders are stucked.

Solution: use packages [charlatan](https://cran.r-project.org/package=wakefield) or [wakefield](https://cran.r-project.org/package=charlatan) to simulate data according to 
the *signed contract*. 

> we were no more stuck while arguing with the other department.



---
class: middle, inverse, center

# Use case: RTE 


---
# Antares

> a Monte-Carlo software for power systems analysis: 
A New Tool for Adequacy Reports and Economic Simulations.

* Generation / Load balance studies
* Economic assessment of Generation projects
* Economic assessment of Transmission projects

.left-column[
  ## The need
]

.right-column[

A tool to generate <span class="bigtext">validated datasets</span> based on data provided by ENTSOE (*european network of transmission system operators for electricity*).

This datasets will be used to play various Monte-carlo scenarii with `antares`.

]


---
# Antares

> a Monte-Carlo software for power systems analysis: 
A New Tool for Adequacy Reports and Economic Simulations.

* Generation / Load balance studies
* Economic assessment of Generation projects
* Economic assessment of Transmission projects

.left-column[
  ## The need
  ## Difficulties
]

.right-column[

* Each country is sending *highly staked* information. Treatments must be specific for each country.
* Aggregation instructions differ for each country.
* Some measures are wrong and must be corrected

]


---
# Antares

> a Monte-Carlo software for power systems analysis: 
A New Tool for Adequacy Reports and Economic Simulations.

* Generation / Load balance studies
* Economic assessment of Generation projects
* Economic assessment of Transmission projects

.left-column[
  ## The need
  ## Difficulties
  ## Open source
]

.right-column[

* Project is beeing hosted on github: https://github.com/rte-antares-rpackage/antaDraft
* Later on CRAN
* roughly 50% complete
* Datasets will be open-sourced also!

]


---

```{r echo=FALSE, results='hide', message=FALSE}
library(antaDraft)
library(magrittr)
library(timevis)
library(DT)
library(glue)
library(purrr)
library(data.table)
library(flextable)
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggiraph)
library(UpSetR)
library(ggalt)
library(randomForest)

theme_ <- theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

load_path <- "/Users/davidgohel/Github/dat.entsoe/load"
```

## read raw datasets

```{r}
load_data <- anta_load_read(data_dir = load_path) 
```

```{r}
sample_n(load_data, size = 3) %>% regulartable() %>% theme_vanilla() %>% autofit()
```

> Which countries did not send data from 2015 to 2016?

```{r}
load_data %>% filter(observed) %>% select(DateTime, country) %>% group_by(country) %>% 
    summarise(start = min(DateTime, na.rm = TRUE), end = max(DateTime, na.rm = TRUE) ) %>% 
    filter( year(start) >= 2015 ) %>% 
  regulartable() %>% theme_vanilla() %>% autofit()
```


---
## Raw validations

```
rules:

- expr: observed==TRUE
  name: IS_OBS
- expr: is.finite(TotalLoadValue)
  name: IS_FINITE
- expr: sign(TotalLoadValue)>0
  name: IS_POS
```


```{r message=FALSE}
load_data_v <- antaDraft::augment_validation(load_data)
```

```{r echo=FALSE}
load_data_v %>% sample_n(size = 3) %>% regulartable() %>% theme_vanilla() %>% autofit()
```


---
## Raw validations

> Quality control summary to be sent to data provider

```{r eval=FALSE}
qc <- load_data_v %>% qualcon()
dir.create(path = "qc/qc_raw", recursive = TRUE)
render_quality(qc, dir = "qc/qc_raw")
```

```{r echo=FALSE}
htmltools::tags$iframe(src = "qc/qc_raw/FRANCE_[IS_OBS].html", style="width:80%;height:500px;border:0;")
```


---
## Datasets creation (aggregate messy data...)


```{r}
aggregated_db <- aggregate_with_rules(load_data_v)
head(sample_n(aggregated_db, size = 10))
```

---
## Datasets creation - How?

```
ITALY:
  CTY:
    - IT
  CTA:
    - IT
  BZN:
    - IT_CNOR
    - IT_CSUD
    - IT_NORD
    - IT_SARD
    - IT_SICI
    - IT_SUD

GERMANY:
  CTY:
    - DE
  CTA:
    - DE_TenneT_GER
    - DE_TransnetBW
    - DE_Amprion
    - DE_50HzT
  BZN:
    - DE_AT_LU
    - "!CTY|AUSTRIA"
    - "!CTY|LUXEMBOURG"

```



---
## Datasets validations

.left-column[
  ## yaml files
]

.right-column[

Simple rules to run row by row:

```
rules:

- expr: is.finite(CTY)
  name: CTY_NA
- expr: is.finite(CTA)
  name: CTA_NA
- expr: is.finite(BZN)
  name: BZN_NA

- expr: ((CTY - lag(CTY)) / CTY) < .5 & ((lag(CTY) - CTY) / lag(CTY)) < .5
  name: CTY_LAG_LT_50
- expr: ((CTA - lag(CTA)) / CTA) < .5 & ((lag(CTA) - CTA) / lag(CTA)) < .5
  name: CTA_LAG_LT_50
- expr: ((BZN - lag(BZN)) / BZN)  < .5 & ((lag(BZN) - BZN) / lag(BZN)) < .5
  name: BZN_LAG_LT_50
```

]


---
## Datasets validations

.left-column[
  ## yaml files
]

.right-column[

Some rules need to be filtered (false positives)

```
- drop:
    - CTY_IS_POS
    - CTY_CTA_EQUAL
    - CTY_BZN_EQUAL
    - CTY_CTA_DIFF_LT_05
    - CTY_BZN_DIFF_LT_05
    - CTY_CTA_DIFF_LT_10
    - CTY_BZN_DIFF_LT_10
    - CTY_LAG_LT_50
  when_not:
    - CTY_NA

- drop:
    - CTA_BZN_DIFF_LT_05
    - CTA_BZN_EQUAL
  when_not:
    - CTA_BZN_DIFF_LT_10
```

]


---
## Datasets validations

.left-column[
  ## yaml files
  ## Validations
]

.right-column[

```{r}
aggregated_db <- augment_validation(aggregated_db)
head(sample_n(aggregated_db, size = 10))
```

]




---
## Datasets validations

.left-column[
  ## yaml files
  ## Validations
]

.right-column[

> made with package `UpSetR`

```{r message=FALSE, echo=FALSE}
features <- as.matrix( aggregated_db[, attr(aggregated_db, "validators")] ) 
features <- !features
mode(features) <- "integer"
datafeat <- features[aggregated_db$country %in% "FRANCE" & year(aggregated_db$DateTime) == 2015 ,] %>% as.data.frame()
upset(datafeat, nsets = 7)
```

]



---
## Datasets validations

.left-column[
  ## yaml files
  ## Validations
]

.right-column[

> This country is not playing the game ;)

```{r message=FALSE, echo=FALSE}
datafeat <- features[aggregated_db$country %in% "LUXEMBOURG",] %>% as.data.frame()
upset(datafeat, nsets = 7)
```

]



---
## Datasets corrections

.left-column[
  ## yaml file
]

.right-column[

> Corrections expressed by an expert

```
-
  replace: CTA
  use: CTY
  when_false:
    - CTA_NA
  when_true:
    - BZN_NA
    - CTY_NA
    - CTY_BZN_EQUAL

-
  replace: CTY
  use: CTA
  country:
  when_false:
    - CTY_NA
  when_true:
    - BZN_NA
    - CTA_NA
    - CTA_BZN_EQUAL
```


]



---
## Datasets corrections

.left-column[
  ## yaml file
  ## corrections
]

.right-column[

```{r}
corrected_db <- aggregated_db %>% 
  data_correct_with_rules() %>% 
  augment_process_summary() 
```

]





---
## Datasets corrections

.left-column[
  ## yaml file
  ## corrections
]

.right-column[

```{r echo=FALSE}
dat_analysis <- corrected_db %>% 
  group_by(country, summary ) %>%
  tally() %>% ungroup() %>%
  tidyr::complete(country, summary, fill = list(n=0) ) 

dat_analysis %>%
  tidyr::spread(summary, n) %>%
  regulartable() %>%
  theme_booktabs() %>% 
  fontsize(size = 13, part = "header") %>% 
  bold(j = 1) %>% 
  padding(padding.left = 5, padding.right = 5, part = "all") %>% 
  border(j = 2:4, border.left = fp_border(), part = "all") %>% 
  autofit()
```

]





---
## More corrections

Here we are trying to complete invalid data with predictions from a model (random forest)


```{r message=FALSE}
rt_by_country <- readRDS("rf/rt_by_country.RDS")
```

```{r message=FALSE, echo=FALSE}
rt_by_country %>% mutate( rsq = map_dbl(model, function(x) mean(x$rsq) ) ) %>% 
  select(country, rsq) %>% regulartable() %>% theme_booktabs()
```


---
## More corrections


```{r message=FALSE}
corrected_db = complete_with_model(corrected_db, rt_by_country)
```

```{r echo=FALSE}
dat_analysis <- corrected_db %>% 
  group_by(country, summary ) %>%
  tally() %>% ungroup() %>%
  tidyr::complete(country, summary, fill = list(n=0) ) 

dat_analysis %>%
  tidyr::spread(summary, n) %>%
  regulartable() %>%
  theme_booktabs() %>% 
  border(j = 2:5, border.left = fp_border(), part = "all") %>% 
  autofit()
```


---

<p class="sing">Thank you</p>


Any questions?

